Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-11-16 paper/introduction.tex
updated intro

Region-based memory management is a good fit for such programs.
A region is a block of memory (consisting of one or more contiguous
range of memory locations, permitting a region to grow) that is used to realize
a data-structure with a well-defined lifetime. 
Once a region is created, individual objects may be dynamically allocated within
the region over time. Finally, the entire region is typically freed in one shot.
In this example, the output to be constructed for each time window $t$ can be
implemented using a separate region.

@@

The goal of this work is a \emph{memory-safe} region-based memory management technique
that supports dynamic regions (that are first-class objects) and higher-order functions.
Though our focus, in this paper, is on memory-safety, our approach offers other side-benefits.
When a producer transfers a data-structure to a consumer, in a shared-memory context , our approach
provides an ownership discipline that ensures that the producer cannot subsequently modify
the data-structure. When the same transfer happens in a distributed setting, the system guarantees
that the data-structure is self-contained and can be safely transmitted across address spaces.
We refer to dynamic regions that provide these guarantees as \emph{transferable} regions.

The key memory safety property we wish to ensure is that there are no dangling references:
i.e., a reference to an object that has been freed.
A key invariant and restriction that we utilize to achieve this is that no object $o_1$ in one dynamic
region $D_1$ can contain a pointer to an object $o_2$ in another dynamic region $D_2$.
(In cases where this is too restrictive, the user has to resort to (deep) copying $o_2$ to $D_1$.
This is conceptually quite reasonable given the perspective that $D_1$ is a self-contained data-structure.
Furthermore, this is quite analogous, from a performance perspective, to the copying that happens
when a garbage collector promotes an object from a lower generation to a higher generation.
From this perspective, a region may be seen as a user-controlled generation.)

@@

\begin{itemize}
  \item We present \name, a \csharp-like typed object-oriented language
  that eschews garbage collection in favour of programmer-managed
  memory regions . \name extends its core language, which includes
  \emph{lambdas} (higher-order functions) and \emph{generics}
  (parametric polymorphism), with constructs to create, manage and
  destroy static and dynamic memory regions. Dynamic regions are
  first-class values in \name; they manifest as objects of type
  \C{Region}, and are treated on par with other objects. \name is
  particularly suited for dataflow programs, such as~\naiad,
  whose memory behavior exhibits
  certain characteristics that admits efficient region-based memory management.
  Experiments demonstrate that   \naiad~\cite{naiad} dataflow program using programmer-managed
  memory regions outperform their GC counterparts by a margin of upto
  59\%.

  \item \name is equipped with a region type system that statically
  guarantees safety of all memory accesses in a well-typed program,
  subject to some conditions that can be checked efficiently at
  run-time.
  The overhead of checking these conditions is less than
  \GK{x\%} in our experiments with \naiad workloads.

>>>>

Region-based memory management is a good fit for such programs.
A region is a block of memory (consisting of one or more contiguous
range of memory locations, permitting a region to grow).
% that is used to realize a data-structure with a well-defined lifetime. 
Once a region is created, individual objects may be dynamically allocated within
the region over time. Finally, the entire region is freed in one shot.
In this example, the output to be constructed for each time window $t$ can be
implemented using a separate region.

@@

The goal of this work is a \emph{memory-safe} region-based memory management technique
that supports dynamic regions (that are first-class objects) and higher-order functions.
Though we focus, in this paper, on memory-safety, our approach offers other benefits.
When a producer transfers a data-structure to a consumer, in a shared-memory context , our approach
provides an ownership discipline that ensures that the producer cannot subsequently modify
the data-structure. When the same transfer happens in a distributed setting, the system guarantees
that the data-structure is self-contained and can be safely transmitted across address spaces.
We refer to dynamic regions that provide these guarantees as \emph{transferable} regions.

The key memory safety property we wish to ensure is that there are no dangling references:
i.e., a reference to an object that has been freed.
A key invariant and restriction that we utilize to achieve this is that no object $o_1$ in one dynamic
region $D_1$ can contain a pointer to an object $o_2$ in another dynamic region $D_2$.
In cases where this is too restrictive, the user has to resort to (deep) copying $o_2$ to $D_1$.
This is conceptually quite reasonable given the perspective that $D_1$ is a self-contained data-structure.
Furthermore, this is quite analogous, from a performance perspective, to the copying that happens
when a garbage collector promotes an object from a lower generation to a higher generation.
From this perspective, a region may be seen as a user-controlled generation.

@@

\begin{itemize}
  \item We present \name, a \csharp-like typed object-oriented language
  that eschews garbage collection in favour of programmer-managed
  memory regions . \name extends its core language, which includes
  \emph{lambdas} (higher-order functions) and \emph{generics}
  (parametric polymorphism), with constructs to create, manage and
  destroy static and dynamic memory regions. Dynamic regions are
  first-class values in \name.
  % Experiments demonstrate that   \naiad~\cite{naiad} dataflow program using programmer-managed
  % memory regions outperform their GC counterparts by a margin of upto
  % 59\%.

  \item \name is equipped with a region type system that statically
  guarantees safety of all memory accesses in a well-typed program,
  provided that the region usage correctly follows a simple discipline.
  The correctness of region usage is checked efficiently at run-time.
  The overhead of checking these conditions is less than
  \GK{x\%} in our experiments with \naiad workloads.
